Grammar-Based Concept Alignment for Domain-Specific Machine Translation

### 2021-06-30 paper/paper.tex
improved captions and references

2 studia VERB 0 root
  1 Lei PRON 2 nsubj
  4 costanza NOUN 2 obl
    3 con ADP 4 case
    \end{verbatim}
  \end{subfigure}
  \caption{The graphical, simplified CoNNL-U and sorted rose tree representation of a pair of UD sentences. With the default criteria, the resulting alignments are: $\langle$\textit{She studies consistently, Lei studia con costanza}$\rangle$, $\langle$\textit{studies, studia}$\rangle$, $\langle$\textit{she, lei}$\rangle$ and $\langle$\textit{consistently, con costanza}$\rangle$.}
\end{figure*}

@@

\paragraph{Known translation divergence}
Parallel texts often present significant, systematic cross-linguistic grammatical distinctions. 
When this is the case, it is often straightforward to define alignment criteria based on recognizing the corresponding patterns.
While many distinctions of this kind are specific to particular language pairs or even stylistical, some of them occur independently of what languages are involved and do not depend on idiomatic usage nor aspectual, discourse, domain or word knowledge.   
Drawing inspiration from \cite{dorr-1994-machine}, we refer to them as \textit{translation divergences} and handle some of the most common ones explicitly. 
For instance, \textit{categorial divergences} occur when the POS tag of a word in the source language changes in its translation.
A ubiquitous example of is that of adverbial modifiers (with the UD label \texttt{advmod}) translated as prepositional phrases (with the UD label \texttt{obl}, for \textit{oblique}), such as in the English-Italian pair $\langle$\textit{She studies \textbf{consistently}, Lei studia \textbf{con costanza}}$\rangle$.
\textit{Structural divergences}, where a direct object in one language is rendered as an oblique in the other, as in the English-Swedish pair $\langle$\textit{I told \textbf{him}, Jag berättade \textbf{för honom}.}$\rangle$), are also frequently encountered.

@@

\begin{table*}[h]
  \centering
  \small
  \begin{tabular}{r|c|c|c|c}
  \multicolumn{1}{l|}{\textbf{}} & \multicolumn{2}{c|}{\textbf{\begin{tabular}[c]{@{}c@{}}PUD (100 sentences)\end{tabular}}} & \multicolumn{2}{c}{\textbf{course plans}}                                                                                                                      \\ \hline
  \multicolumn{1}{l|}{}          & \textbf{en-it}                               & \textbf{en-sv}                               & \textbf{\begin{tabular}[c]{@{}c@{}}DMI (881 sentences)\end{tabular}} & \textbf{\begin{tabular}[c]{@{}c@{}}CSE (539 sentences)\end{tabular}} \\ \hline
  \textbf{distinct alignments}        & 1197                                         & 1325                                         & 1823                                                                           & 1950                                                                           \\
  \textbf{correct}                & 916 (77\%)                                   & 1112 (85\%)                                  & 1205 (66\%)                                                                    & 1296 (66\%)                                                                    \\
  \textbf{usable in MT}                      & 880 (74\%)                                   & 1099 (84\%)                                  & 1157 (63\%)                                                                    & 1248 (64\%)                                                                    \\
  \end{tabular}
  \caption{Comparison between the extraction of alignments of any size from manually annotated PUD treebanks and from automatically parsed sentences from the course plans corpora.}
  \label{raw_fast}
  \end{table*}

>>>>

2 studia VERB 0 root
  1 Lei PRON 2 nsubj
  4 costanza NOUN 2 obl
    3 con ADP 4 case
    \end{verbatim}
  \end{subfigure}
  \caption{The graphical, simplified CoNNL-U and sorted rose tree representation of a pair of UD sentences. With the default criteria, which among other things allow for matching adverbial with adjectival modifiers, the resulting alignments are: $\langle$\textit{She studies consistently, Lei studia con costanza}$\rangle$, $\langle$\textit{studies, studia}$\rangle$, $\langle$\textit{she, lei}$\rangle$ and $\langle$\textit{consistently, con costanza}$\rangle$.}
  \label{ex}
\end{figure*}

@@

\paragraph{Known translation divergence}
Parallel texts often present significant, systematic cross-linguistic grammatical distinctions. 
When this is the case, it is often straightforward to define alignment criteria based on recognizing the corresponding patterns.
While many distinctions of this kind are specific to particular language pairs or even stylistical, some of them occur independently of what languages are involved and do not depend on idiomatic usage nor aspectual, discourse, domain or word knowledge.   
Drawing inspiration from \cite{dorr-1994-machine}, we refer to them as \textit{translation divergences} and handle some of the most common ones explicitly. 
For instance, \textit{categorial divergences} occur when the POS tag of a word in the source language changes in its translation.
An ubiquitous example of is that of adverbial modifiers (with the UD label \texttt{advmod}) translated as prepositional phrases (with the UD label \texttt{obl}, for \textit{oblique}), such as in the English-Italian pair $\langle$\textit{She studies \textbf{consistently}, Lei studia \textbf{con costanza}}$\rangle$ (see Figure \ref{ex}).
\textit{Structural divergences}, where a direct object in one language is rendered as an oblique in the other, as in the English-Swedish pair $\langle$\textit{I told \textbf{him}, Jag berättade \textbf{för honom}.}$\rangle$), are also frequently encountered.

@@

\begin{table*}[h]
  \centering
  \small
  \begin{tabular}{r|c|c|c|c}
  \multicolumn{1}{l|}{\textbf{}} & \multicolumn{2}{c|}{\textbf{\begin{tabular}[c]{@{}c@{}}PUD (100 sentences)\end{tabular}}} & \multicolumn{2}{c}{\textbf{course plans}}                                                                                                                      \\ \hline
  \multicolumn{1}{l|}{}          & \textbf{en-it}                               & \textbf{en-sv}                               & \textbf{\begin{tabular}[c]{@{}c@{}}DMI (881 sentences)\end{tabular}} & \textbf{\begin{tabular}[c]{@{}c@{}}CSE (539 sentences)\end{tabular}} \\ \hline
  \textbf{distinct alignments}        & 1197                                         & 1325                                         & 1823                                                                           & 1950                                                                           \\
  \textbf{correct}                & 916 (77\%)                                   & 1112 (85\%)                                  & 1205 (66\%)                                                                    & 1296 (66\%)                                                                    \\
  \textbf{usable in MT}                      & 880 (74\%)                                   & 1099 (84\%)                                  & 1157 (63\%)                                                                    & 1248 (64\%)                                                                    \\
  \end{tabular}
  \caption{Comparison between the grammar-based extraction of alignments of any size from manually annotated PUD treebanks and from automatically parsed sentences from the course plans corpora.}
  \label{raw_fast}
  \end{table*}
